---
title: "Prevalence of Diarrhea for FBD"
author: "Michael Yousuf, Arie Havelaar, Getachew Teshome"
format: html
editor: visual
bibliography: "https://api.citedrive.com/bib/8bc1f4de-a9ad-44c9-8d6b-094c43271b3d/references.bib?x=eyJpZCI6ICI4YmMxZjRkZS1hOWFkLTQ0YzktOGQ2Yi0wOTRjNDMyNzFiM2QiLCAidXNlciI6ICIxMTIzMCIsICJzaWduYXR1cmUiOiAiM2ZiNjYyMWFmZDlkMjNkMzA2Njg2ZTEzOGJhMmU3ZjZiNGQ5YmQwMmUzOGI4MWIyZjIwYTdlYjJjZWNmOGI2ZSJ9= .bib"
---

## Prevalence of Diarrhea for \<= 5 age.

If we want to carry out a study to understand the prevalence of diarrhea among the children under the age of 5 in East Hararge here is what i think it will represent the sample size of our interest. A previous study carried out by @alebel2018 stated that the prevalence of diarrhea among under-five children in Ethiopia was 22%, 22.5% @mengistie2013 for eastern part of Ethiopia and 23% @teferasolomon2022 in the study of finding the prevalence of diarrhea in the under five children in rural Dire Dawa city. Using the sample size formula in @sharma2019 the final number of participant in this study with the expected prevalence of 22% at 95% of CI and 5% margin of error will be...

```{r}

library(epiR)
```

```{r}
#calculate for the 
epi.sssimpleestb(N = NA, Py = 0.22 , epsilon = 0.25, error = "relative", 
   se = 0.95 , sp = 0.95 , nfractional = FALSE, conf.level = 0.95)
```

```{r}
293 + 29 # considering 10% dropout of study participants
```

The `echo: false` option disables the printing of code (only output is displayed).

## Final report

For the expected prevalence of 22%, the required sample size is 322 for the margin of error or absolute precision of +5 or -5 in estimating the prevalence with 95% confidence and considering a potential loss/attrition of 10%. with this sample size, the anticipated 95% CI is (17%, 27%). this sample size is calculated using epiR (R statistical software)
